Correlation tests

### 2025-04-21 analysis/manuscript.qmd
Arxiv submission

Transformative agreements are a much-discussed library licensing model for transitioning subscription-based journal publishing to full open access.
Although these agreements can vary considerably, they are mainly aimed at hybrid journal bundles, enabling authors from participating institutions to publish open access in these journals while providing reading access to the entire portfolio [@Borrego_2020; @Hinchliffe_2019].
Following their initial proposal in 2015 [@Schimmer_2015], the number of transformative agreements has grown substantially, resulting in increased open access publishing in hybrid journals [@Jahn_2025; @Rothfritz_2024; @McCabe_2024].
Prior to these agreements, only a few articles in hybrid journals were openly accessible [@Bj_rk_2012; @Laakso_2016; @Piwowar_2018], except for some European countries that had dedicated funding policies and the SCOAP$^3$ consortium for high-energy physics journals [@Pinfield_2015; @Robinson_Garcia_2020; @Huang_2020; @Kohls_2018].
Coordination efforts through initiatives such as OA2020 and Efficiency and Standards for Article Charges (ESAC) have further streamlined workflows between library consortia and publishers [@ifla; @Geschuhn_2017].
By April 2025, the ESAC Transformative Agreement Registry, the primary community-driven evidence source, listed approximately 1,300 agreements.[^esac_registry]

@@

Following @Jahn_2025, the starting point was a unified dataset of several safeguarded JCT snapshots[^jct_data].
JCT journal data were enriched with ISSN variants linked to ISSN-L according to the ISSN agency. 
To identify hybrid journals, a comprehensive exclusion of fully open access journals was performed using multiple journal lists including the Directory of Open Access Journals (DOAJ).
The resulting hybrid journal data were made available via hoaddata and were used to independently determine the publication volume for each database.

@@

The author confirms the sole responsibility for the following: Conceptualization, Methodology, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Writing - Review & Editing, Visualization.

@@

The author declares no competing interests.

@@

This work was supported by the Federal Ministry of Education and Research of Germany (BMBF) under grants 16WIK2301E / 16WIK2101F.

The proprietary bibliometric data from Scopus and Web of Science were provided by the German Competence Network for Bibliometrics under BMBF grant 16WIK2101A.

>>>>

Transformative agreements are a much-discussed library licensing model for transitioning subscription-based journal publishing to full open access.
Although these agreements can vary considerably, they are mainly aimed at hybrid journal bundles, enabling authors from participating institutions to publish open access in these journals while providing reading access to the entire portfolio [@Borrego_2020; @Hinchliffe_2019].
Following their initial proposal in 2015 [@Schimmer_2015], the number of transformative agreements has grown substantially, resulting in increased open access publishing in hybrid journals [@Jahn_2025; @Rothfritz_2024; @McCabe_2024].
Prior to these agreements, only a few articles in hybrid journals were openly accessible [@Bj_rk_2012; @Laakso_2016; @Piwowar_2018], except for some European countries that had dedicated funding policies and the SCOAP$^3$ consortium for high-energy physics journals [@Pinfield_2015; @Robinson_Garcia_2020; @Huang_2020_elife; @Kohls_2018].
Coordination efforts through initiatives such as OA2020 and Efficiency and Standards for Article Charges (ESAC) have further streamlined workflows between library consortia and publishers [@ifla; @Geschuhn_2017].
By April 2025, the ESAC Transformative Agreement Registry, the primary community-driven evidence source, listed approximately 1,300 agreements.[^esac_registry]

@@

Following @Jahn_2025, the starting point was a unified dataset of several safeguarded JCT snapshots[^jct_data].
JCT journal data were enriched with ISSN variants linked to ISSN-L according to the ISSN agency. 
To identify hybrid journals, a comprehensive exclusion of fully open access journals was performed using multiple journal lists including the Directory of Open Access Journals (DOAJ).
The resulting hybrid journal data comprising more tahn 13,000 hybrid journals were made available via hoaddata and were used to independently determine the publication volume for each database.

@@

The author confirms the sole responsibility for the following: Conceptualisation, Methodology, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Writing - Review & Editing, Visualization.

@@

The author has participated in working groups of OA2020, ESAC and the Barcelona Declaration as part of his professional responsibilities.
No funding was received from these initiatives for the present study, and they had no influence on its design, data collection, analysis, or interpretation of results.

@@

This work was supported by the Federal Ministry of Education and Research of Germany (BMBF) under grants 16WIK2301E / 16WIK2101F.
The proprietary bibliometric data from Scopus and Web of Science were provided by the German Competence Network for Bibliometrics under BMBF grant 16WIK2101A.

### 2025-04-21 analysis/manuscript.tex
Arxiv submission

\section*{Abstract}
{This study compares open metadata from hoaddata, an openly available dataset based on Crossref, OpenAlex and the cOAlition S Journal Checker Tool, with proprietary bibliometric databases Scopus and Web of Science to estimate the impact of transformative agreements on hybrid open access publishing. Analysing over 13,000 hybrid journals between 2019-2023, the research found substantial growth in open access due to these agreements, although most articles remain paywalled. The results were consistent across all three data sources, showing strong correlations in country-level metrics despite differences in journal coverage and metadata availability. By 2023, transformative agreements enabled the majority of hybrid open access articles, with particularly high adoption in European countries. The analysis revealed strong alignment between first and corresponding authorship when measuring agreement uptake by publisher and country. This comparative approach supports the use of open metadata for large-scale hybrid open access studies, while using multiple data sources together provides a more robust understanding of hybrid open access adoption than any single database can offer, overcoming individual limitations in coverage and metadata quality.}

@@

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, enabling authors from participating
institutions to publish open access in these journals while providing
reading access to the entire portfolio (Borrego et al., 2021;
Hinchliffe, 2019). Following their initial proposal in 2015 (Schimmer et
al., 2015), the number of transformative agreements has grown
substantially, resulting in increased open access publishing in hybrid
journals (Jahn, 2025; McCabe \& Mueller-Langer, 2024; Rothfritz et al.,
2024). Prior to these agreements, only a few articles in hybrid journals
were openly accessible (Björk, 2012; Laakso \& Björk, 2016; Piwowar et
al., 2018), except for some European countries that had dedicated
funding policies and the SCOAP\(^3\) consortium for high-energy physics
journals (Huang et al., 2020; Kohls \& Mele, 2018; Pinfield, 2015;
Robinson-Garcia et al., 2020). Coordination efforts through initiatives
such as OA2020 and Efficiency and Standards for Article Charges (ESAC)
have further streamlined workflows between library consortia and
publishers (Campbell et al., 2022; Geschuhn \& Stone, 2017). By April
2025, the ESAC Transformative Agreement Registry, the primary
community-driven evidence source, listed approximately 1,300
agreements.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}

@@

While promising for general bibliometric research, applying open
metadata to transformative agreements presents distinct challenges due
to their targeted journal coverage and complex funding attribution. With
invoicing data largely unavailable, previous studies have relied on
publication fee pricing lists or expenditures, and first and/or
corresponding author affiliations as proxies to attribute funding to
institutions (Butler et al., 2023; Haustein et al., 2024; Zhang et al.,
2022). Recent work has extended this approach and attributed open access
articles to transformative agreements by using data from the cOAlition S
Journal Checker Tool, which links journals and participating
institutions to transformative agreements included in the ESAC registry
(Jahn, 2025; Jonge et al., 2025). Yet, no systematic comparison exists
of how open versus proprietary data sources perform when analysing the
same hybrid journals covered by transformative agreements to identify
comparative strengths and shortcomings of data sources.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L
according to the ISSN agency. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists including the Directory of Open Access
Journals (DOAJ). The resulting hybrid journal data were made available
via hoaddata and were used to independently determine the publication
volume for each database.

@@

Based on these compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain ROR-IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included
matching of both journals and participating institutions according to
the JCT. The matching also considered the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (Jonge et al., 2025), applied to publications
funded by the Dutch Research Council (NWO) and validated against
internal invoicing data, confirmed that such matching can accurately
identify most articles under transformative agreements.

@@

The findings were consistent across the investigated open data source
hoaddata, derived from Crossref and OpenAlex, and the established
proprietary bibliometric databases Scopus and Web of Science. This
aligns with previous studies (Akbaritabar et al., 2024; Alperin et al.,
2024) and rankings (Eck et al., 2024). Overall, the results show strong
correlations by country affiliation, which supports the use open
metadata for large-scale analyses of hybrid open access. However, the
observed differences in journal coverage and metadata availability
warrant further discussion, affecting not only open data sources but
also proprietary databases when used in isolation.

@@

The database comparison revealed important discrepancies in open access
evidence. Crucially, not all publishers share CC licence metadata via
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely
because of parsing errors on the publisher's website. This resulted in
fewer open access articles being recorded in Scopus and Web of Science.
Further inconsistencies emerged between Scopus and Web of Science,
despite both relying on Unpaywall: Web of Science erroneously labels
Elsevier's delayed open access as hybrid, whereas Scopus correctly
categorises it. Notably, Scopus and Crossref showed greater alignment
than that observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the use of a curated list
to identify hybrid open access journals, rather than depending solely on
article-level tags. Given these variations, which may compromise
comparability, research and monitoring exercises should avoid relying on
a single source. Instead, the selection of open access evidence should
be cross-verified using multiple sources and snapshots that can be used
to track changes in the data over time (Huang et al., 2020).
Incorporating expert-curated journal lists can also help prevent
misclassification based on business models (Visser et al., 2021).

It is important to note that the study's estimates of articles from
institutions involved in transformative agreements are approximations
because of a lack of access to invoice data, which is not usually shared
by library consortia and publishers. Furthermore, the open data sources
for transformative agreements, including the cOAlition S Journal Checker
Tool and the ESAC registry, are based on voluntary effort, crowd-sourced
from various consortia. However, recent validation using Dutch research
information demonstrates the reliability of such an open approach for
the assessment of articles under transformative agreements (Jonge et
al., 2025). With enhanced open metadata compliance attributable to
evolving standards and initiatives in support of negotiations with
publishers, particularly through the ESAC initiative and Barcelona
Declaration on Open Research Information, the situation is likely to
improve. In conclusion, this study has shown that open metadata are
well-suited for analysing the transition from subscription-based journal
publishing to full open access. But using them in conjunction with
proprietary databases provides a more robust understanding that
overcomes individual shortcomings in coverage and metadata quality.

@@

The author confirms the sole responsibility for the following:
Conceptualization, Methodology, Formal analysis, Investigation, Data
Curation, Writing - Original Draft, Writing - Review \& Editing,
Visualization.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Achterberg_2023}
Achterberg, I., \& Jahn, N. (2023). Introducing the hybrid open access
dashboard (HOAD). \emph{cOAlition S Blog}.
\url{https://www.coalition-s.org/blog/introducing-the-hybrid-open-access-dashboard-hoad/}

@@

@@

\bibitem[\citeproctext]{ref-eck_2024_6bx73}
Eck, N. J. van, Waltman, L., \& Neijssel, M. (2024). \emph{{Launch of
the CWTS Leiden Ranking Open Edition 2024}}. Leiden Madtrics.
\url{https://doi.org/10.59350/r512t-r8h93}

@@

@@

\bibitem[\citeproctext]{ref-de_Jonge_2025}
Jonge, H. de, Kramer, B., \& Sondervan, J. (2025). \emph{Tracking
transformative agreements through open metadata: Method and validation
using {Dutch Research Council NWO} funded papers}. MetaArXiv.
\url{https://doi.org/10.31222/osf.io/tz6be_v1}

@@

\bibitem[\citeproctext]{ref-Robinson_Garcia_2020}
Robinson-Garcia, N., Costas, R., \& Leeuwen, T. N. van. (2020). Open
access uptake by universities worldwide. \emph{{PeerJ}}, \emph{8},
e9410. \url{https://doi.org/10.7717/peerj.9410}

@@

\bibitem[\citeproctext]{ref-Visser_2021}
Visser, M., Eck, N. J. van, \& Waltman, L. (2021). Large-scale
comparison of bibliographic data sources: {Scopus, Web of Science,
Dimensions, Crossref, and Microsoft Academic}. \emph{Quantitative
Science Studies}, \emph{2}(1), 20--41.
\url{https://doi.org/10.1162/qss_a_00112}

>>>>

\section*{Abstract}
{This study compares open metadata from hoaddata, an openly available dataset based on Crossref, OpenAlex and the cOAlition S Journal Checker Tool, with proprietary bibliometric databases Scopus and Web of Science to estimate the impact of transformative agreements on hybrid open access publishing. Analysing over 13,000 hybrid journals between 2019-2023, the research found substantial growth in open access due to these agreements, although most articles remain paywalled. The results were consistent across all three data sources, showing strong correlations in country-level metrics despite differences in journal coverage and metadata availability. By 2023, transformative agreements enabled the majority of open access in hybrid journals, with particularly high adoption in European countries. The analysis revealed strong alignment between first and corresponding authorship when measuring agreement uptake by publisher and country. This comparative approach supports the use of open metadata for large-scale hybrid open access studies, while using multiple data sources together provides a more robust understanding of hybrid open access adoption than any single database can offer, overcoming individual limitations in coverage and metadata quality.}

@@

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, enabling authors from participating
institutions to publish open access in these journals while providing
reading access to the entire portfolio (Borrego et al., 2021;
Hinchliffe, 2019). Following their initial proposal in 2015 (Schimmer et
al., 2015), the number of transformative agreements has grown
substantially, resulting in increased open access publishing in hybrid
journals (Jahn, 2025; McCabe \& Mueller-Langer, 2024; Rothfritz et al.,
2024). Prior to these agreements, only a few articles in hybrid journals
were openly accessible (Björk, 2012; Laakso \& Björk, 2016; Piwowar et
al., 2018), except for some European countries that had dedicated
funding policies and the SCOAP\(^3\) consortium for high-energy physics
journals (C.-K. (Karl). Huang et al., 2020; Kohls \& Mele, 2018;
Pinfield, 2015; Robinson-Garcia et al., 2020). Coordination efforts
through initiatives such as OA2020 and Efficiency and Standards for
Article Charges (ESAC) have further streamlined workflows between
library consortia and publishers (Campbell et al., 2022; Geschuhn \&
Stone, 2017). By April 2025, the ESAC Transformative Agreement Registry,
the primary community-driven evidence source, listed approximately 1,300
agreements.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}

@@

While promising for general bibliometric research, applying open
metadata to transformative agreements presents distinct challenges due
to their targeted journal coverage and complex funding attribution. With
invoicing data largely unavailable, previous studies have relied on
publication fee pricing lists or expenditures, and first and/or
corresponding author affiliations as proxies to attribute funding to
institutions (Butler et al., 2023; Haustein et al., 2024; Zhang et al.,
2022). Recent work has extended this approach and attributed open access
articles to transformative agreements by using data from the cOAlition S
Journal Checker Tool, which links journals and participating
institutions to transformative agreements included in the ESAC registry
(de Jonge et al., 2025; Jahn, 2025). Yet, no systematic comparison
exists of how open versus proprietary data sources perform when
analysing the same hybrid journals covered by transformative agreements
to identify comparative strengths and shortcomings of data sources.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L
according to the ISSN agency. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists including the Directory of Open Access
Journals (DOAJ). The resulting hybrid journal data comprising more tahn
13,000 hybrid journals were made available via hoaddata and were used to
independently determine the publication volume for each database.

@@

Based on these compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain ROR-IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included
matching of both journals and participating institutions according to
the JCT. The matching also considered the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (de Jonge et al., 2025), applied to
publications funded by the Dutch Research Council (NWO) and validated
against internal invoicing data, confirmed that such matching can
accurately identify most articles under transformative agreements.

@@

The findings were consistent across the investigated open data source
hoaddata, derived from Crossref and OpenAlex, and the established
proprietary bibliometric databases Scopus and Web of Science. This
aligns with previous studies (Akbaritabar et al., 2024; Alperin et al.,
2024) and rankings (van Eck et al., 2024). Overall, the results show
strong correlations by country affiliation, which supports the use open
metadata for large-scale analyses of hybrid open access. However, the
observed differences in journal coverage and metadata availability
warrant further discussion, affecting not only open data sources but
also proprietary databases when used in isolation.

@@

The database comparison revealed important discrepancies in open access
evidence. Crucially, not all publishers share CC licence metadata via
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely
because of parsing errors on the publisher's website. This resulted in
fewer open access articles being recorded in Scopus and Web of Science.
Further inconsistencies emerged between Scopus and Web of Science,
despite both relying on Unpaywall: Web of Science erroneously labels
Elsevier's delayed open access as hybrid, whereas Scopus correctly
categorises it. Notably, Scopus and Crossref showed greater alignment
than that observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the use of a curated list
to identify hybrid open access journals, rather than depending solely on
article-level tags. Given these variations, which may compromise
comparability, research and monitoring exercises should avoid relying on
a single source. Instead, the selection of open access evidence should
be cross-verified using multiple sources and snapshots that can be used
to track changes in the data over time (C.-K. Huang et al., 2020).
Incorporating expert-curated journal lists can also help prevent
misclassification based on business models (Visser et al., 2021).

It is important to note that the study's estimates of articles from
institutions involved in transformative agreements are approximations
because of a lack of access to invoice data, which is not usually shared
by library consortia and publishers. Furthermore, the open data sources
for transformative agreements, including the cOAlition S Journal Checker
Tool and the ESAC registry, are based on voluntary effort, crowd-sourced
from various consortia. However, recent validation using Dutch research
information demonstrates the reliability of such an open approach for
the assessment of articles under transformative agreements (de Jonge et
al., 2025). With enhanced open metadata compliance attributable to
evolving standards and initiatives in support of negotiations with
publishers, particularly through the ESAC initiative and Barcelona
Declaration on Open Research Information, the situation is likely to
improve. In conclusion, this study has shown that open metadata are
well-suited for analysing the transition from subscription-based journal
publishing to full open access. But using them in conjunction with
proprietary databases provides a more robust understanding that
overcomes individual shortcomings in coverage and metadata quality.

@@

The author confirms the sole responsibility for the following:
Conceptualisation, Methodology, Formal analysis, Investigation, Data
Curation, Writing - Original Draft, Writing - Review \& Editing,
Visualization.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Achterberg_2023}
Achterberg, I., \& Jahn, N. (2023). \emph{Introducing the hybrid open
access dashboard (HOAD)}. {cOAlition S}.
\url{https://www.coalition-s.org/blog/introducing-the-hybrid-open-access-dashboard-hoad/}

@@

\bibitem[\citeproctext]{ref-de_Jonge_2025}
de Jonge, H., Kramer, B., \& Sondervan, J. (2025). \emph{Tracking
transformative agreements through open metadata: Method and validation
using {Dutch Research Council NWO} funded papers}. MetaArXiv.
\url{https://doi.org/10.31222/osf.io/tz6be_v1}

@@

@@

\bibitem[\citeproctext]{ref-Huang_2020_elife}
Huang, C.-K. (Karl)., Neylon, C., Hosking, R., Montgomery, L., Wilson,
K. S., Ozaygen, A., \& Brookes-Kenworthy, C. (2020). Evaluating the
impact of open access policies on research institutions. \emph{eLife},
\emph{9}. \url{https://doi.org/10.7554/elife.57067}

@@

@@

\bibitem[\citeproctext]{ref-Robinson_Garcia_2020}
Robinson-Garcia, N., Costas, R., \& van Leeuwen, T. N. (2020). Open
access uptake by universities worldwide. \emph{{PeerJ}}, \emph{8},
e9410. \url{https://doi.org/10.7717/peerj.9410}

@@

\bibitem[\citeproctext]{ref-eck_2024_6bx73}
van Eck, N. J., Waltman, L., \& Neijssel, M. (2024). \emph{{Launch of
the CWTS Leiden Ranking Open Edition 2024}}. Leiden Madtrics.
\url{https://doi.org/10.59350/r512t-r8h93}

\bibitem[\citeproctext]{ref-Visser_2021}
Visser, M., van Eck, N. J., \& Waltman, L. (2021). Large-scale
comparison of bibliographic data sources: {Scopus, Web of Science,
Dimensions, Crossref, and Microsoft Academic}. \emph{Quantitative
Science Studies}, \emph{2}(1), 20--41.
\url{https://doi.org/10.1162/qss_a_00112}
